Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-08-02 paper.tex
more

We can now write a simple proof that shows the union of \(\top\) and \(\top\)
is equivalent to the Boolean type, that is {\tt (⊤ + ⊤) ↔ 𝔹}.
Alternatively, since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we
name our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 : (⊤ + ⊤) ↔ 𝔹
1+1=2 = ?
\end{code}
\end{minipage}
\end{center}
To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
given Boolean. This works out quite simply since the type {\tt ⊤ + ⊤} only has
two possible terms, {\tt (inl *)} and {\tt (inr *)}, which we can associate with
the two terms, {\tt True} and {\tt False}, in the type {\tt 𝔹} and vice-versa.
Thus, we can write the proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 = <> (f , g)
  where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
        g = rec𝔹 (⊤ +' ⊤) (inl *) (inr *)
\end{code}
\end{minipage}
\end{center}
For our purposes, we use the constructor {\tt <>} for constructing a term of
the bijection type. For simplicity, we can define a bijection as a conjunction
of two logical implications:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _↔_ (A B : Set) : Set where
  <>_ : (A → B) × (B → A) → (A ↔ B)
\end{code}
\end{minipage}
\end{center}
To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
equality by logical bijection). We go over the three different forms of
equivalence in the following section.

@@

\subsection*{Equality}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality that is generally more familiar than the others.
This form of equality is called {\it definitional equality} \cite{}. Simply put,
definitional equality is equality through definition, which requires no proof.
For example, we know that for any two natural numbers {\tt m} and {\tt n},
{\tt m `plus` n} is equal to the sum of {\tt m} and {\tt n} by definition of
{\tt `plus`}.

We have already briefly mentoined the second form of equality, judgmental
equality \cite{}, which is equality by logical bijection. For example, if we
have the propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we
can derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
{\tt B} are judgmentally equivalent. A proof of judgmental equality requires a
proof similar to the one we have presented in the previous section for the
proposition {\tt (⊤ + ⊤) ↔ 𝔹}.

The final form of equivalence is quite a bit more complex than the previous two
and requires a special kind of proof. Propositional equality is equality through
an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
definition of the identity type:
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
%% -- "PROPOSITIONAL EQUALITY"
%% -- For a proposition, C, on a proof of x ≡ y for
%% -- any x and y, if we can show that it holds for
%% -- (refl x), then C holds for all proofs, p : x ≡ y.
%% -- "STRICT EQUALITY"
%% -- For a proposition, C, on a reflexive equality (x ≡ x),
%% -- for any x, if we can show that it holds for (refl x), then
%% -- C holds for all other reflexive equalities p : x ≡ x.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _≡_ {A : Set} : (a b : A) → Set where
  refl : (a : A) → (a ≡ a)
\end{code}
\end{minipage}
\end{center}
Out of all the types we have introduced, the identity type is the first type
parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
introduced types parameterized by other types. Types paramterized by terms are
classified as an {\em indexed type} \cite{}. Essentially, if we think of types
as shapes (i.e., we can think of them as circles containing points representing
their terms), indexed types change their shape depending on which terms are
passed to them. For example, it is not the case that {\tt y ≡ x} automatically
comes from a proof of {\tt x ≡ y}. This requires a proof!

@@

rec≡ : {A : Set} {x y : A} (C : A → Set) →
       (p : x ≡ y) → C x → C y
rec≡ {A} {x} {y} C = 
  J (λ {x} {y} _ → C x → C y)
    (λ x → λ z → z)
    {x} {y} 
\end{code}
\end{minipage}
\end{center}

>>>>

We can now write a simple proof that shows the union of \(\top\) and \(\top\)
is equivalent to the Boolean type, that is {\tt (⊤ + ⊤) ↔ 𝔹}.
Alternatively, since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we
name our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 : (⊤ + ⊤) ↔ 𝔹
1+1=2 = ?
\end{code}
\end{minipage}
\end{center}
To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
given Boolean. This works out quite simply since the type {\tt ⊤ + ⊤} only has
two possible terms, {\tt (inl *)} and {\tt (inr *)}, which we can associate with
the two terms, {\tt True} and {\tt False}, in the type {\tt 𝔹} and vice-versa.
Thus, we can write the proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 = <> (f , g)
  where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
        g = rec𝔹 (⊤ +' ⊤) (inl *) (inr *)
\end{code}
\end{minipage}
\end{center}
For our purposes, we use the constructor {\tt <>} for constructing a term of
the bijection type. For simplicity, we can define a bijection as a conjunction
of two logical implications:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _↔_ (A B : Set) : Set where
  <>_ : (A → B) × (B → A) → (A ↔ B)
\end{code}
\end{minipage}
\end{center}
To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
equality by logical bijection) \cite{}. We go over the three different forms of
equivalence in the following section.

@@

\subsection*{Equality}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality is much simpler than the others. This form of
equality is called {\it definitional equality} \cite{}. Simply put, definitional
equality is equality through definition, which requires no proof. For example,
we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.

We have already briefly mentioned the second form of equality, judgmental
equality, which is equality by logical bijection. For example, if we have the
propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we can
derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
{\tt B} are judgmentally equivalent. A proof of judgmental equality requires a
proof similar to the one we have presented in the previous section for the
proposition {\tt (⊤ + ⊤) ↔ 𝔹}.

The final form of equivalence is quite a bit more complex than the previous two
and requires a special kind of proof. Propositional equality is equality through
an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
definition of the identity type:
%% say something that this is the HoTT version of identity
%% possibly also point out that the discrepancy between the two
%% -- "PROPOSITIONAL EQUALITY"
%% -- For a proposition, C, on a proof of x ≡ y for
%% -- any x and y, if we can show that it holds for
%% -- (refl x), then C holds for all proofs, p : x ≡ y.
%% -- "STRICT EQUALITY"
%% -- For a proposition, C, on a reflexive equality (x ≡ x),
%% -- for any x, if we can show that it holds for (refl x), then
%% -- C holds for all other reflexive equalities p : x ≡ x.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _≡_ {A : Set} : (a b : A) → Set where
  refl : (a : A) → (a ≡ a)
\end{code}
\end{minipage}
\end{center}
Out of all the types we have introduced, the identity type is the first type
parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
introduced types parameterized by other types. Types parameterized by terms are
classified as an {\em indexed types} \cite{}. Essentially, if we think of types
as shapes (i.e., we can think of them as circles containing points representing
their terms), indexed types change their shape depending on which terms are
passed to them. For example, it is not the case that {\tt y ≡ x} automatically
comes from a proof of {\tt x ≡ y}. This requires a proof of the symmetry of
identity types.

@@

rec≡ : {A : Set} {x y : A} (C : A → Set) →
       (p : x ≡ y) → C x → C y
rec≡ {A} {x} {y} C = 
  J (λ {x} {y} _ → C x → C y)
    (λ x → λ z → z)
    {x} {y} 
\end{code}
\end{minipage}
\end{center}
The two induction principles of the identity type (viz. {\tt J} and {\tt K})
refer to the two ways in which an identity type can be used in a proof, where
{\tt J} is more general. This simply means that {\tt J} can be used for a wider
variety of proofs concerning identity types.

At first glance, the difference between the two induction principles is clear
when looking at the {\tt C} parameter in both. In {\tt J}, {\tt C} is
parameterized with two possibly distinct terms (viz. {\tt x} and {\tt y}),
whereas {\tt K} only mentions one, {\tt x}. This distinction is what allows
{\tt J} to be used for {\em special} identity proofs, that is, in proofs
for identity types of more complex forms of equality (i.e., propositional
equality).

As an example, let's prove the symmetry of identity types using {\tt J}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J ? ?
\end{code}
\end{minipage}
\end{center}
Now, we must fill in both instances of {\tt ?} in the above proof. Similar to
previous proofs, the first of these is our motive and is simply a restatement of
our type statement. In these forms of proofs, we commonly annotate parameterized
identity types with their given equality (viz. {\tt x=y}):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J (λ {x y} x=y → y ≡ x) ?
\end{code}
\end{minipage}
\end{center}
Now, the last {\tt ?} must be filled in with a proof that our motive holds in
the case that {\tt y} is {\tt x}, that is, we must prove that
{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
proof is the following:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym_J = J (λ {x y} x=y → y ≡ x) refl
\end{code}
\end{minipage}
\end{center}
